---
title: "Reproducible Research: Peer Assessment 1"
author: Jakub Duda
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.path="figure/assign1-")
library(dplyr)
library(lubridate)
```

## Loading and preprocessing the data

```{r}
data<-read.csv('activity.csv')
head(data)
```

## What is mean total number of steps taken per day?

Let's plot the histogram of average daily steps:

```{r}
data_sum<-data %>% na.omit() %>% group_by(date) %>% summarize(suma=sum(steps))
hist(data_sum[[2]],main="Histogram of Total Daily Steps",xlab="Number of Steps",breaks=10)
```

The mean number of steps is

```{r}
mean(data_sum$suma)
```

and the median is

```{r}
median(data_sum$suma)
```

## What is the average daily activity pattern?

```{r}
data_mean_interval<-data %>% na.omit() %>% group_by(interval) %>% summarize(mean=mean(steps))
plot((1:288)*5,data_mean_interval[,2][[1]],type="l",col="blue",
     xlab="Minutes Since Day Start",ylab="Average Number of Steps",
     main="AverageSteps Per 5 Minute Intervals")
```

The maximum number of steps on average is in the interval:

```{r}
idx <- which.max(data_mean_interval$mean)
max_int <- data_mean_interval[idx,]$interval
print(max_int)
```

## Imputing missing values

The number of missing values is:

```{r}
sum(is.na(data$steps))
```

We will replace the missing number of steps by the corresponding average for the given 5-minute interval.

```{r}
data_fixed<-data
for (i in 1:nrow(data_fixed))
{
  if (is.na(data_fixed[i,]$steps))
  {
    data_fixed[i,]$steps <-
      data_mean_interval[
        data_mean_interval$interval==data_fixed[i,]$interval,]$mean
  }
}
```

Let's plot a histogram of the data with imputed values:

```{r}
data_sum_fixed<- data_fixed %>% na.omit()  %>% group_by(date) %>%  summarize(suma=sum(steps))
hist(data_sum_fixed[[2]],
     main="Histogram of Total Daily Steps (Data Fixed)",
     xlab="Number of Steps",
     breaks=10)
```

Let's compare the means and medians of the original and imputed data:

```{r}
typeofval<-c("mean","median","mean","median")
datatype<-c("original","original","imputed","imputed")
vals<-c(mean(data_sum$suma),
        median(data_sum$suma),
        mean(data_sum_fixed$suma),
        median(data_sum_fixed$suma))
df<-data.frame(type_of_analytic=typeofval,data=datatype,value=vals)
df
```

The mean remains the same after imputing, while the median for the imputed data increases slightly.

## Are there differences in activity patterns between weekdays and weekends?

We add a weekday/weekend factor and compare the patterns on the imputed data:

```{r}
data_fixed$weekday <- weekdays(ymd(as.character(data_fixed$date)))
data_fixed$dayType="NA"
for (i in 1:nrow(data_fixed))
{
  if ((data_fixed[i,]$weekday=="Saturday") | (data_fixed[i,]$weekday=="Sunday"))
  {
    data_fixed[i,]$dayType="weekend"
  } else
  {
    data_fixed[i,]$dayType="weekday"
  } 
}
```

Finally, we compare weekday and weekend activity:
```{r}
par(mfrow=c(2,1))

data_weekday<-data_fixed %>% filter(dayType=="weekday") %>% group_by(interval) %>% summarize(mean=mean(steps))
plot((1:288)*5,data_weekday[,2][[1]],type="l",col="blue",
     xlab="Minutes Since Day Start",ylab="Average Number of Steps",
     main="Weekday Average Steps")

data_weekend<-data_fixed %>% filter(dayType=="weekend") %>% group_by(interval) %>% summarize(mean=mean(steps))
plot((1:288)*5,data_weekend[,2][[1]],type="l",col="blue",
     xlab="Minutes Since Day Start",ylab="Average Number of Steps",
     main="Weekend Average Steps")
```

The profiles of weekday and weekend activity differ. Weekends correspond to lower peak activity, while activity is more uniformly spread during wake-time hours.
